---
title: "Connecting R to a Database"
---

## Learning Objectives

By the end of this chapter, students will be able to:

- **Explain** the two-layer DBI design: a standard interface package and database-specific driver packages.
- **Open and close** a PostgreSQL connection using `dbConnect` and `dbDisconnect`, including with `on.exit` for safe cleanup.
- **Retrieve** query results as R data frames using `dbGetQuery` and `dbListTables`/`dbListFields` for schema exploration.
- **Visualize** database query results using `ggplot2` within an R script or notebook.
- **Write** parameterized queries using the `params` argument to prevent SQL injection.
- **Execute** DDL and DML statements that return no rows using `dbExecute`.

```{r}
#| include: false
source("_common.R")

con <- dbConnect(
  Postgres(),
  host     = "localhost",
  port     = as.integer(Sys.getenv("PGPORT", unset = "5432")),
  dbname   = "nycflights",
  user     = "postgres",
  password = Sys.getenv("PGPASS", unset = "postgres")
)
```

Throughout this book, SQL code blocks have been executing against a live PostgreSQL database without much explanation of how that connection actually works. The connection is established by R using the `DBI` package, and every query result comes back to R as a data frame. This chapter makes that mechanism explicit, showing you how to connect to a database from R, retrieve query results, visualize them with `ggplot2`, and write queries safely using parameterization.

These skills matter outside of Quarto documents. Most real data science work happens in R scripts and notebooks where you need to pull data from a database, analyze it in R, and produce outputs. Understanding how to drive that connection directly gives you full control over the process.

## The DBI Ecosystem

R's database connectivity is organized around a two-layer design. The `DBI` package defines a standard interface: a set of functions with consistent names and behavior regardless of which database you are connecting to. Underneath DBI, **driver packages** implement that interface for specific databases.

| Driver package | Database |
|---|---|
| `RPostgres` | PostgreSQL |
| `RSQLite` | SQLite |
| `RMariaDB` | MySQL / MariaDB |
| `bigrquery` | Google BigQuery |
| `odbc` | Any ODBC-compatible database |

Because DBI provides a uniform interface, the code for connecting to and querying a database looks almost identical regardless of which database you use. Swapping from SQLite to PostgreSQL means changing one line (the driver), not rewriting your analysis.

## Establishing a Connection

`dbConnect()` creates a connection to the database. It takes a driver object as its first argument, followed by connection parameters specific to that driver.

```{r}
#| label: connect
#| eval: false
library(DBI)
library(RPostgres)

con <- dbConnect(
  Postgres(),
  host     = "localhost",
  port     = 5433,
  dbname   = "nycflights",
  user     = "postgres",
  password = "postgres"
)
```

The object `con` represents an open connection to the database. It is passed to every subsequent DBI function that needs to communicate with the database.

Before writing any queries, it is useful to explore what the database contains:

```{r}
#| label: list-tables
dbListTables(con)
```

```{r}
#| label: list-fields
dbListFields(con, "flights")
```

`dbListTables()` returns the names of all tables in the connected database. `dbListFields()` returns the column names for a specific table.

## Querying Data

`dbGetQuery()` sends a SQL query to the database and returns the result as a data frame. It is the most common function for analytical work: you write the SQL, and R gives you back a data frame you can work with immediately.

```{r}
#| label: get-query
delays_by_carrier <- dbGetQuery(con, "
  SELECT
      f.carrier,
      a.name,
      COUNT(*)                               AS flights,
      ROUND(AVG(f.dep_delay)::numeric, 1)    AS mean_delay,
      ROUND(STDDEV(f.dep_delay)::numeric, 1) AS stddev_delay
  FROM flights AS f
  JOIN airlines AS a ON f.carrier = a.carrier
  WHERE f.dep_delay IS NOT NULL
  GROUP BY f.carrier, a.name
  ORDER BY mean_delay DESC
")

delays_by_carrier
```

The result is a standard R data frame. Every tool you use for data manipulation in R, including `dplyr`, `data.table`, and base R, works on the result immediately.

```{r}
#| label: class-check
class(delays_by_carrier)
```

## Visualizing Query Results

Because `dbGetQuery()` returns a data frame, you can pipe the result directly into `ggplot2`. The following sections build three plots using the data retrieved in the previous section and a few additional queries.

### Distribution of Departure Delays

A histogram reveals the shape of the delay distribution discussed in the previous chapter:

```{r}
#| label: hist-delays
#| warning: false
library(ggplot2)

delays_raw <- dbGetQuery(con, "
  SELECT dep_delay
  FROM flights
  WHERE dep_delay IS NOT NULL
    AND dep_delay BETWEEN -30 AND 180
")

ggplot(delays_raw, aes(x = dep_delay)) +
  geom_histogram(binwidth = 5, fill = "#0165fc", color = "white") +
  labs(
    title = "Distribution of Departure Delays",
    subtitle = "LGA flights, 2013 — trimmed to −30 to +180 minutes",
    x = "Departure Delay (minutes)",
    y = "Number of Flights"
  ) +
  theme_minimal()
```

The right skew is unmistakable. The bulk of flights cluster just below zero, and a long tail of severe delays extends to the right — the same pattern the mean-vs-median comparison in the previous chapter quantified numerically.

### Departure Delay vs. Arrival Delay

The strong correlation between departure and arrival delay is easier to see as a scatterplot than as a single number:

```{r}
#| label: scatter-delays
#| warning: false
delay_pairs <- dbGetQuery(con, "
  SELECT dep_delay, arr_delay
  FROM flights
  WHERE dep_delay IS NOT NULL
    AND arr_delay IS NOT NULL
    AND dep_delay BETWEEN -30 AND 120
    AND arr_delay BETWEEN -60 AND 120
")

ggplot(delay_pairs, aes(x = dep_delay, y = arr_delay)) +
  geom_point(alpha = 0.05, color = "#0165fc") +
  geom_smooth(method = "lm", color = "firebrick", se = FALSE) +
  labs(
    title = "Departure Delay vs. Arrival Delay",
    subtitle = "Each point is one flight; regression line overlaid",
    x = "Departure Delay (minutes)",
    y = "Arrival Delay (minutes)"
  ) +
  theme_minimal()
```

The regression line runs nearly through the origin with a slope close to 1, consistent with the coefficients computed in SQL in the previous chapter. The transparency (`alpha = 0.05`) reveals where the mass of flights lies despite over 100,000 overlapping points.

### Mean Delay by Carrier

A horizontal bar chart makes carrier comparisons easy to read:

```{r}
#| label: bar-carrier
ggplot(delays_by_carrier, aes(x = mean_delay, y = reorder(name, mean_delay))) +
  geom_col(fill = "#0165fc") +
  labs(
    title = "Mean Departure Delay by Carrier",
    subtitle = "LGA flights, 2013",
    x = "Mean Departure Delay (minutes)",
    y = NULL
  ) +
  theme_minimal()
```

`reorder(name, mean_delay)` sorts the bars by delay value rather than alphabetically, making the ranking immediately visible.

## Parameterized Queries

When a query needs to incorporate a value from R, the instinct is to paste it directly into the query string:

```{r}
#| label: unsafe-query
#| eval: false
# Do not do this
carrier_code <- "AA"
dbGetQuery(con, paste0("SELECT * FROM flights WHERE carrier = '", carrier_code, "'"))
```

This is unsafe. If `carrier_code` came from user input and contained something like `'; DROP TABLE flights; --`, the database would execute that injected SQL. The correct approach is a **parameterized query**, which passes values separately from the query text so the database treats them as data, never as SQL.

DBI supports parameterized queries through the `params` argument. Placeholders in the query string are written as `$1`, `$2`, and so on (for PostgreSQL):

```{r}
#| label: parameterized-query
carrier_code <- "AA"

aa_flights <- dbGetQuery(
  con,
  "SELECT flight, dep_delay, arr_delay, dest
   FROM flights
   WHERE carrier = $1
     AND dep_delay IS NOT NULL
   ORDER BY dep_delay DESC",
  params = list(carrier_code)
)

head(aa_flights)
```

The value of `carrier_code` is sent to the database separately. No matter what string is in `carrier_code`, it is treated as a literal value to compare against the `carrier` column, not as SQL to execute. Multiple parameters follow the same pattern: add more `$n` placeholders and extend the `params` list.

::: {.callout-tip}
Always use parameterized queries whenever a query incorporates a value that originates from outside your own code, including user input, configuration files, or other data sources. The `params` argument is the right tool in DBI.
:::

## Non-Returning Statements

`dbGetQuery()` is for queries that return rows. For statements that modify data — `INSERT`, `UPDATE`, `DELETE`, `CREATE TABLE` — use `dbExecute()`. It sends the statement to the database and returns the number of rows affected rather than a result set.

```{r}
#| label: db-execute
#| eval: false
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS carrier_summary AS
  SELECT
      carrier,
      COUNT(*)                            AS total_flights,
      ROUND(AVG(dep_delay)::numeric, 1)  AS avg_delay
  FROM flights
  WHERE dep_delay IS NOT NULL
  GROUP BY carrier
")
```

Like `dbGetQuery()`, `dbExecute()` accepts a `params` argument for safe value substitution.

## Closing the Connection

A database connection holds resources on both the R side and the database server side. Close it explicitly when you are done:

```{r}
#| label: disconnect
#| eval: false
dbDisconnect(con)
```

In a script that opens a connection at the top and closes it at the bottom, it is good practice to use `on.exit()` to ensure the connection closes even if an error occurs partway through:

```{r}
#| label: on-exit
#| eval: false
con <- dbConnect(Postgres(), host = "localhost", port = 5433,
                 dbname = "nycflights", user = "postgres", password = "postgres")
on.exit(dbDisconnect(con), add = TRUE)

# ... rest of your analysis ...
```

`on.exit(..., add = TRUE)` schedules `dbDisconnect(con)` to run when the surrounding function or script exits, regardless of whether it exits normally or with an error.

## Exercises

### Reflection

1. What is the relationship between the `DBI` package and driver packages like `RPostgres`? Why is this two-layer design useful?

2. Explain what is wrong with building a query string using `paste0()` to insert a user-supplied value. What does a parameterized query do differently, and why does it prevent SQL injection?

3. When would you use `dbExecute()` instead of `dbGetQuery()`? Give an example of a SQL statement appropriate for each.

### Coding

4. Connect to the `nycflights` database and use `dbListFields()` to inspect the columns of the `weather` table. Then write a `dbGetQuery()` call that retrieves the average temperature (`temp`) and average wind speed (`wind_speed`) for each month. Store the result in a data frame and print it.

5. Using `dbGetQuery()`, retrieve the number of flights per destination (`dest`) from the `flights` table. Then use `ggplot2` to create a bar chart of the top 15 destinations by flight count.

6. Write a parameterized query that accepts a minimum number of Oscar nominations as a parameter and returns matching actors from the `actors` database. (You will need to open a second connection to the `actors` database.) Test it with at least two different threshold values.

7. Retrieve daily average departure delay from `flights` (aggregate by `month` and `day`). Plot the result as a line chart with `ggplot2`, using `month` and `day` to construct a date on the x-axis. Describe the seasonal pattern you observe.
