---
title: "Indexes"
---

## Learning Objectives

By the end of this chapter, students will be able to:

- **Explain** what a B-tree index is and describe the trade-off between read performance and write and storage overhead.
- **Read** a query execution plan produced by `EXPLAIN ANALYZE` and interpret the cost, row, and timing annotations.
- **Identify** the three main scan strategies (Sequential Scan, Bitmap Heap Scan, Index Scan) and explain when the planner chooses each.
- **Create** a single-column index and verify with `EXPLAIN ANALYZE` that the planner uses it.
- **Create** a multi-column index and explain how column order affects which queries benefit from it.
- **Write** a partial index that covers only a subset of rows and identify when it is preferable to a full index.
- **Use** the `INCLUDE` clause to create a covering index that satisfies a query entirely from the index without touching the heap.
- **Use** `CREATE INDEX CONCURRENTLY` to build an index without locking the table for writes.
- **Diagnose** cases where an index exists but the planner ignores it, and explain the common causes.

```{r}
#| include: false
source("_common.R")

con <- dbConnect(
  Postgres(),
  host     = "localhost",
  port     = as.integer(Sys.getenv("PGPORT", unset = "5432")),
  dbname   = "nycflights",
  user     = "postgres",
  password = Sys.getenv("PGPASS", unset = "postgres")
)

dbExecute(con, "DROP INDEX IF EXISTS idx_flights_dep_delay")
dbExecute(con, "DROP INDEX IF EXISTS idx_flights_month_carrier")
dbExecute(con, "DROP INDEX IF EXISTS idx_flights_dep_delay_covering")
dbExecute(con, "DROP INDEX IF EXISTS idx_flights_delayed")
```

Every query discussed in the design chapters so far has assumed tables small enough that performance was not a concern. In practice, tables grow. A `flights` table with 100,000 rows behaves quite differently from one with 100 million, and a query that returns in milliseconds today can take seconds or minutes as the data volume increases. Indexes are the primary tool for keeping read performance acceptable as tables scale.

This chapter uses the `nycflights` database. The `flights` table has 104,662 rows and only one index out of the box: a composite primary key. That gives a clean baseline for observing exactly what each new index changes.

## What an Index Is

An index is a separate data structure that the database maintains alongside a table. It stores a copy of one or more columns in a sorted, searchable form, together with a pointer back to the full row in the table (called the **heap**). When a query filters on an indexed column, PostgreSQL can traverse the index to find the matching rows directly, rather than reading every row in the table.

The trade-off is straightforward: indexes make reads faster and writes slower. Every `INSERT`, `UPDATE`, or `DELETE` that touches an indexed column must also update the index. Indexes also consume disk space. A thoughtfully chosen index speeds up the queries that matter most; an excessive number of indexes on a write-heavy table imposes overhead that outweighs any benefit.

PostgreSQL supports several index types. By far the most common is the **B-tree** (balanced tree), which is the default when you write `CREATE INDEX` without specifying a type. A B-tree organizes values in sorted order, which makes it efficient for equality checks (`=`), range queries (`<`, `>`, `BETWEEN`), and sorting. Unless you have a specific reason to choose otherwise, B-tree is the right choice.

## Reading Query Plans

Before creating any index, it is worth understanding how to read the query plan that PostgreSQL produces. The `EXPLAIN` command shows the plan without executing the query; `EXPLAIN ANALYZE` executes it and adds actual timing and row counts alongside the estimates.

```{r}
#| label: explain-no-index
writeLines(dbGetQuery(con,
  "EXPLAIN ANALYZE SELECT * FROM flights WHERE dep_delay > 120"
)[[1]])
```

The plan output is a tree of **nodes**, each representing one operation. Reading the cost annotation on a node:

```
Seq Scan on flights  (cost=0.00..3178.28 rows=2778 width=81)
                     (actual time=0.019..11.192 rows=2791 loops=1)
```

- **cost=0.00..3178.28** — estimated work to return the first row (startup) and all rows (total). These are relative units the planner uses to compare strategies, not milliseconds.
- **rows=2778** — estimated row count. The actual count follows in the ANALYZE output.
- **width=81** — average bytes per output row.
- **actual time=0.019..11.192** — measured wall-clock milliseconds to first row and last row.
- **loops=1** — how many times this node executed. Nodes inside a nested loop execute once per outer row.

The `Rows Removed by Filter: 101871` line shows how many rows the scan read and discarded. For this query, PostgreSQL examined all 104,662 rows to find 2,791 that matched, taking about 11 milliseconds.

### The Three Scan Strategies

PostgreSQL chooses among three scan strategies depending on how selective the filter is relative to the table size:

| Strategy | When chosen | Characteristic |
|---|---|---|
| **Sequential Scan** | Large fraction of the table | Reads all pages linearly |
| **Bitmap Heap Scan** | Moderate fraction | Builds a page bitmap, then fetches only those pages |
| **Index Scan** | Very few rows | Traverses index leaf by leaf, fetching each heap row immediately |

A sequential scan is not always wrong. For filters that match a large percentage of a table, reading the table linearly is faster than the random I/O of an index lookup. The planner chooses based on estimated cost.

## Creating a Basic Index

The `dep_delay > 120` query above identified severely delayed flights: 2,791 rows, or about 2.7 percent of the table. That selectivity is a good candidate for an index.

```{r}
#| label: create-basic-index
#| include: false
dbExecute(con, "CREATE INDEX idx_flights_dep_delay ON flights(dep_delay)")
```

With the index in place, the plan changes:

```{r}
#| label: explain-with-index
writeLines(dbGetQuery(con,
  "EXPLAIN ANALYZE SELECT * FROM flights WHERE dep_delay > 120"
)[[1]])
```

The planner switched to a **Bitmap Heap Scan**. It first ran a `Bitmap Index Scan` to traverse the new index and build an in-memory bitmap of pages containing matching rows, then fetched only those pages from the heap. Execution time dropped from about 11 milliseconds to about 2 — a roughly five-fold improvement.

For a more selective filter — flights delayed over five hours — the same index is even more effective:

```{r}
#| label: explain-highly-selective
writeLines(dbGetQuery(con,
  "EXPLAIN ANALYZE SELECT * FROM flights WHERE dep_delay > 300"
)[[1]])
```

Only 228 rows match. With so few results, PostgreSQL uses a pure **Index Scan** rather than the bitmap approach, fetching each matching heap row directly from the index. Execution time falls under a millisecond.

The shift from Bitmap Heap Scan to Index Scan reflects the planner's logic: when very few pages need to be fetched, the overhead of building a bitmap is not worth it. Random I/O directly from the index is faster.

```{r}
#| label: drop-basic-index
#| include: false
dbExecute(con, "DROP INDEX idx_flights_dep_delay")
```

## Multi-Column Indexes

When queries frequently filter on two or more columns together, a multi-column index can be more efficient than separate single-column indexes.

Without any additional index, filtering on both month and carrier requires a full sequential scan:

```{r}
#| label: explain-multi-before
writeLines(dbGetQuery(con,
  "EXPLAIN ANALYZE SELECT * FROM flights WHERE month = 7 AND carrier = 'UA'"
)[[1]])
```

652 rows match. Creating a composite index on both columns eliminates the full scan:

```{r}
#| label: create-multi-index
#| include: false
dbExecute(con, "CREATE INDEX idx_flights_month_carrier ON flights(month, carrier)")
```

```{r}
#| label: explain-multi-after
writeLines(dbGetQuery(con,
  "EXPLAIN ANALYZE SELECT * FROM flights WHERE month = 7 AND carrier = 'UA'"
)[[1]])
```

Execution time drops from about 11 milliseconds to well under 1 millisecond.

### Column Order and the Leading-Column Rule

The order of columns in a multi-column index determines which queries can use it. An index on `(month, carrier)` is organized first by month, then by carrier within each month. A query that filters on `month` alone can use this index, because month is the leading column. A query that filters on `carrier` alone cannot use the index efficiently — carrier values are interleaved across all months, so the index provides no useful ordering.

```{r}
#| label: explain-non-leading
writeLines(dbGetQuery(con,
  "EXPLAIN ANALYZE SELECT * FROM flights WHERE carrier = 'UA'"
)[[1]])
```

PostgreSQL does use the index here, but the cost is much higher than the two-column query: it must scan nearly the entire index to collect all United Airlines entries scattered across all 12 months. A dedicated single-column index on `carrier` would outperform this. The general rule is to place the most selective column, or the one that appears in the most queries, first.

```{r}
#| label: drop-multi-index
#| include: false
dbExecute(con, "DROP INDEX idx_flights_month_carrier")
```

## Partial Indexes

A **partial index** covers only the rows that satisfy a `WHERE` clause specified at creation time. It is smaller than a full index, builds faster, and uses less memory.

The `flights` table has 3,153 rows where `dep_delay IS NULL` — cancelled or diverted flights that never received a departure delay measurement. Queries about departure delays almost never want these rows. An index on `dep_delay` that includes them wastes space on values that will never be matched in a filter like `dep_delay > 0`.

```sql
CREATE INDEX idx_flights_delayed ON flights(dep_delay)
WHERE dep_delay > 0;
```

This index covers only the 33,690 rows with a positive delay — about a third of the table. Queries that filter on positive delays use the partial index just as they would a full index, but the index itself is one-third the size.

```{r}
#| label: create-partial-index
#| include: false
dbExecute(con, "CREATE INDEX idx_flights_delayed ON flights(dep_delay) WHERE dep_delay > 0")
```

```{r}
#| label: explain-partial
writeLines(dbGetQuery(con,
  "EXPLAIN ANALYZE SELECT * FROM flights WHERE dep_delay > 120"
)[[1]])
```

The plan shows the partial index in use. The planner recognizes that `dep_delay > 120` is a subset of the partial index's `dep_delay > 0` condition and uses it accordingly. A query that does not satisfy the partial index predicate — such as `WHERE dep_delay IS NULL` — will not use it and will fall back to a sequential scan.

```{r}
#| label: drop-partial-index
#| include: false
dbExecute(con, "DROP INDEX idx_flights_delayed")
```

## Covering Indexes

An **index-only scan** occurs when every column needed by a query is present in the index itself, meaning PostgreSQL never needs to touch the heap at all. The `INCLUDE` clause adds extra columns to the index specifically for this purpose, without making them part of the sort order.

Consider a query that selects only `dep_delay`, `carrier`, and `dest`:

```{r}
#| label: explain-covering-before
writeLines(dbGetQuery(con,
  "EXPLAIN ANALYZE SELECT dep_delay, carrier, dest FROM flights WHERE dep_delay > 120"
)[[1]])
```

Without an index, this is a sequential scan. A standard index on `dep_delay` would enable a bitmap heap scan, but the heap fetch would still be necessary to retrieve `carrier` and `dest`. Adding those columns to the index via `INCLUDE` eliminates the heap fetch entirely:

```{r}
#| label: create-covering-index
#| include: false
dbExecute(con,
  "CREATE INDEX idx_flights_dep_delay_covering ON flights(dep_delay) INCLUDE (carrier, dest)"
)
```

```{r}
#| label: explain-covering-after
writeLines(dbGetQuery(con,
  "EXPLAIN ANALYZE SELECT dep_delay, carrier, dest FROM flights WHERE dep_delay > 120"
)[[1]])
```

The plan now shows `Index Only Scan` with `Heap Fetches: 0`. Every value the query needs is available directly from the index pages, without reading the heap. Execution time drops from about 11 milliseconds to under half a millisecond.

Covering indexes are most valuable for queries that run very frequently on large tables and return only a few columns. The cost is a larger index on disk.

```{r}
#| label: drop-covering-index
#| include: false
dbExecute(con, "DROP INDEX idx_flights_dep_delay_covering")
```

## Expression Indexes

PostgreSQL can index the result of a function or expression rather than a raw column value. This is necessary when queries apply a function to a column in the `WHERE` clause.

A common example is case-insensitive email lookup. A schema that stores email addresses as mixed case cannot use a plain index on `email` to satisfy `WHERE LOWER(email) = 'user@example.com'`, because the index stores the original case. An expression index on `LOWER(email)` solves the problem:

```sql
CREATE INDEX idx_students_email_lower ON students(LOWER(email));
```

A query written as `WHERE LOWER(email) = 'user@example.com'` will now use this index. A query written as `WHERE email = 'User@Example.com'` will not, because the expressions do not match.

The same principle applies to any function used in a filter: `EXTRACT(year FROM created_at)`, `date_trunc('month', created_at)`, `jsonb_extract_path_text(payload, 'event_type')`. If a function appears in your `WHERE` clause and the query is slow, an expression index on that function is likely the fix.

## CREATE INDEX CONCURRENTLY

By default, `CREATE INDEX` acquires a lock that blocks all writes to the table for the duration of the build. On a small table this takes milliseconds. On a large production table it can block inserts and updates for minutes.

`CREATE INDEX CONCURRENTLY` builds the index in the background without blocking writes:

```sql
CREATE INDEX CONCURRENTLY idx_flights_dep_delay ON flights(dep_delay);
```

The trade-off is that it takes longer to build, requires two passes over the table, and cannot be run inside a transaction block. If the build fails partway through, it leaves an invalid index behind that must be dropped manually before retrying.

For any table that receives ongoing writes in a production environment, `CONCURRENTLY` is the safe default.

## When Indexes Are Not Used

An index exists but the planner ignores it in several common situations.

**Low selectivity.** If a filter matches a large fraction of the table, the planner correctly prefers a sequential scan. An index on a boolean column in a table where 95 percent of rows have `is_active = true` will rarely be used for `WHERE is_active = true`, because fetching 95 percent of the table through random I/O is slower than a linear scan.

**Function wrapping.** Applying a function to an indexed column in the `WHERE` clause breaks the match between the query and the index. `WHERE EXTRACT(month FROM time_hour) = 7` cannot use an index on `time_hour`, because the index stores raw timestamps, not extracted months. Either write the filter as a range (`WHERE time_hour >= '2013-07-01' AND time_hour < '2013-08-01'`) or create an expression index.

**Type mismatch.** Comparing a column to a value of a different type forces an implicit cast that may prevent index use. `WHERE flight = '301'` (comparing an integer column to a text literal) can suppress the index, depending on how PostgreSQL resolves the cast. Always match the literal type to the column type.

**Stale statistics.** The planner estimates how many rows a filter will return using statistics collected by `ANALYZE`. If those statistics are far out of date, the planner may underestimate selectivity and choose a sequential scan when an index scan would be faster. Running `ANALYZE table_name` refreshes the statistics.

## Identifying Index Candidates

Not every column deserves an index. A useful mental checklist:

- **Columns that appear in `WHERE` clauses frequently.** The more often a column is filtered, the more benefit an index provides.
- **Join keys.** PostgreSQL uses indexes on join columns to avoid full scans when building hash or nested-loop joins.
- **Foreign key columns.** PostgreSQL does not automatically index foreign keys (unlike some other databases). A missing foreign key index can cause slow cascading deletes and slow joins.
- **High-selectivity columns.** A column with many distinct values — a UUID, an email address, a timestamp — benefits more from an index than a column with few distinct values, like a status flag.
- **Columns used in `ORDER BY`.** If a query both filters and sorts on the same column, the index may satisfy both operations, eliminating a sort step.

Columns that are written to constantly, rarely queried, or very low selectivity are poor candidates. Monitor query performance over time and add indexes in response to observed slow queries rather than preemptively indexing every column.
